### 2018-07-24 chapters/03-translator/gen-app.tex
Finalize first version of the journal paper Still 1.000 words too long

\section{Translation Problems}
There are several potential problems to peform translations between systems that embed semantic information in the input. Those problems vary from simple cases, e.g., a function is not defined in system, to complex cases, e.g., different positioning of branch cuts for multivalued functions. This section will discuss the problems and workarounds.

@@

A similar problem are multiplications, since they are rarely explicitely marked in \LaTeX{} expressions, e.g., scientists using whitespaces to indicate multiplications rather than using \verb|\cdot| or similar symbols. For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication sign (this macro will not be rendered). Since this macro is newly introduced and an automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} uses this macro yet. As a consequence, the translator has some simple rules to perform translations without explicitely marked translations with \verb|\idot|.

@@

All translations are defined by a library that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as placeholders to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in table~\ref{tab:placeholder_ex2}

@@

In some special cases, these placeholders causes trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation}
\verb|diff(f, [x$n])|,
\end{equation}
where $f$ as an algebraic expression or an equation, $x$ the name of the differentiation variable and $n$ for the $n$-th order differentiation. A translation for $\frac{d^2x^2}{dx^2}$ should be like this
\begin{equation}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation}
but would end up as
\begin{equation}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff($1, [$2$$0])|.
\end{equation}

@@

We can solve this issue by using paranthesis in such cases, e.g., 
\begin{equation}
\verb|diff($1, [$2$($0)])|.
\end{equation}

@@

\begin{table}[ht]
	\centering
	\begin{tabular}{ll}
		\hline
		Semantic Macro Entry & Maple Entry \\
		\hline
		\verb|\LegendreP{\nu}@{x}| & \verb|LegendreP($0, $1)| \\
		\verb|X1:\LegendrePX\LegendreP[\mu]{\nu}@{x}| & \verb|LegendreP($1, $0, $2)|\\
		\hline
	\end{tabular}
	\caption{Entry of the translation library example of the Legendre and associated Legendre function of the first kind. The prefix notation \texttt{X<d>:<name>X} defines the translation for \texttt{<name>} with \texttt{<d>}-number of optional arguments.}
	\label{tab:legendreP-lex}
\end{table}

@@

\begin{table}
	\centering
	\begin{tabular}{lll}
	\hline
	\multicolumn{3}{l}{Symbol: \texttt{\textbackslash sin}} \\
	\! & \multicolumn{2}{l}{Feature Set: dlmf-macro} \\
	\! & \! & DLMF: \verb|\sin@@{z}|\\
	\! & \! & DLMF-Link: dlmf.nist.gov/4.14\# E1\\
	\! & \! & Meanings: Sine\\
	\! & \! & Number of Parameters: 0\\
	\! & \! & Number of Variables: 1\\
	\! & \! & Number of Ats: 2\\
	\! & \! & Maple: \verb|sin($0)|\\
	\! & \! & Maple-Link: www.maplesoft.com/support/\\
	\! & \! & \hspace{32pt} help/maple/view.aspx?path=sin\\
	\! & \! & Mathematica: \verb|Sin[$0]|\\
	\! & \! & Mathematica-Link: reference.wolfram.com/\\
	\! & \! & \hspace{32pt} language/ref/Sin.html\\
	\hline
	\end{tabular}
	\caption{The entry of the sine function in the lexicon file.}
	\label{tab:sine-lex-example}
\end{table}

>>>>

\section{Translation Problems}\label{sec:problems}
There are several potential problems to peform translations between systems that embed semantic information in the input. Those problems vary from simple cases, e.g., a function is not defined in system, to complex cases, e.g., different positioning of branch cuts for multivalued functions. This section will discuss the problems and workarounds.

@@

A similar problem are multiplications, since they are rarely explicitely marked in \LaTeX{} expressions, e.g., scientists using whitespaces to indicate multiplications rather than using \verb|\cdot| or similar symbols. For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication symbol (this macro will not be rendered). Since this macro is newly introduced and an automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} uses this macro yet. As a consequence, the translator has some simple rules to perform translations without explicitely marked translations with \verb|\idot|.

Even we only allow the special dialect of \LaTeX{} using semantic macros not all expressions are unambigious. In table~\ref{tab:amb-latex} are four examples of ambiguous expressions. These expressions are unambigious for the \LaTeX{} compiler, since it only considers the very next token for power and subscripts. Our translator following the same rules to solve these issues.

\begin{table}[ht]
\centering
\begin{tabular}{cc}
	\hline
	Ambiguous Input & \LaTeX{} Output\\
	\hline
	\verb|n^m!| & $n^m!$\\
	\verb|a^bc^d| & $a^bc^d$\\
	\verb|x^y^z| & Double superscript error\\
	\verb|x_y_z| & Double subscript error\\
	\hline
\end{tabular}
\caption{Ambiguous \LaTeX{} expressions and how \LaTeX{} displays them.}
\label{tab:amb-latex}
\end{table}

Another more questionable translation decision are alphanumerical expressions. As explained in table~\ref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently, depending on the order of the symbols. The reason is, that an expression such as '$4b$' is usually considered to be a multiplication of $4$ and '$b$', while '$b4$' looks like indexing '$b$' by $4$. While the first example produces two nodes, namely $4$ and '$b$', the second example '$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons. (1) we would assume that the inputs '$4b$' and '$b4$' are mathematically equivalent and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we define the following definitions

\begin{eqnarray*}
\verb|4b| & \overset{\langMaple}{\mapsto} & \verb|4*b|,\\
\verb|b4| & \overset{\langMaple}{\mapsto} & \verb|b*4|,\\
\verb|energy| & \overset{\langMaple}{\mapsto} & \verb|e*n*e*r*g*y|.
\end{eqnarray*}

In general the translator is drafted to solve ambiguous expressions or automatically find a workaround to disambiguate the expression. Only if there is no way to solve the ambiguity with the defined rules, the translation process stops.

@@

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as placeholders to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in table~\ref{tab:placeholder_ex2}.

@@

These placeholders causes trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation*}
\verb|diff(f, [x$n])|,
\end{equation*}
where $f$ as an algebraic expression or an equation, $x$ the name of the differentiation variable and $n$ for the $n$-th order differentiation. A translation for $\frac{d^2x^2}{dx^2}$ should be like this
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation*}
but would end up as
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [xx])|.
\end{equation*}

@@

We can solve this issue by using paranthesis in such cases, e.g., \verb|diff($1, [$2$($0)])|.

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{ll}
		\hline
		Semantic Macro Entry & Maple Entry \\
		\hline
		\verb|\LegendreP{\nu}@{x}| & \verb|LegendreP($0, $1)| \\
		\verb|X1:\LegendrePX\LegendreP[\mu]{\nu}@{x}| & \verb|LegendreP($1, $0, $2)|\\
		\hline
	\end{tabular}
	\caption{Example entries of the Legendre and associated Legendre function in the translation library. The prefix notation \texttt{X<d>:<name>X} defines the translation for \texttt{<name>} with \texttt{<d>}-number of optional arguments.}
	\label{tab:legendreP-lex}
\end{table}

@@

\begin{table}[ht!]
	\centering
	\begin{tabular}{lll}
	\hline
	\multicolumn{3}{l}{Symbol: \texttt{\textbackslash sin}} \\
	\! & \multicolumn{2}{l}{Feature Set: dlmf-macro} \\
	\! & \! & DLMF: \verb|\sin@@{z}|\\
	\! & \! & DLMF-Link: dlmf.nist.gov/4.14\# E1\\
	\! & \! & Meanings: Sine\\
	\! & \! & Number of Parameters: 0\\
	\! & \! & Number of Variables: 1\\
	\! & \! & Number of Ats: 2\\
	\! & \! & Maple: \verb|sin($0)|\\
	\! & \! & Maple-Link: www.maplesoft.com/support/\\
	\! & \! & \hspace{32pt} help/maple/view.aspx?path=sin\\
	\! & \! & Mathematica: \verb|Sin[$0]|\\
	\! & \! & Mathematica-Link: reference.wolfram.com/\\
	\! & \! & \hspace{32pt} language/ref/Sin.html\\
	\hline
	\end{tabular}
	\caption{The entry of the sine function in the lexicon file.}
	\label{tab:sine-lex-example}
\end{table}
